On the properties of discs around accreting brown dwarfs

### 2010-05-19 BD_disc/paper.tex
Shortened Rotation rate section (parameters) and changed inner_temp.ps
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2272 25464f21-6099-402c-b30a-640f1d40d38b

XXXX nathan - please reduce by factor 2 in length!

Data for rotation rates, from periodic variability surveys, are widely
available for a range of different age clusters of TTS. However, fewer
studies exist on the rotation rates of BD mass objects. To select a
feasible range of periods for our model grid we have collated the
results from some recent studies of BD variability. Rotation period
data for $\sigma$ Ori, at an age of $\approx$ 3 Myrs
\citep{mayne_2008}, is studied in \cite{scholz_2004},
where periods are found over the range 5.78$-$74.4 hours ($\approx$
0.24$-$3.1 days) for BD mass objects. In some cases the periodic
variability is irregular and assumed to come from active accretion hot
spots on the BD surface \citep[see][for discussion of variability
causes]{bouvier_1995,herbst_2007}. Infrared excesses
and evidence of accretion are also found to correlate with slower
rotating TTS, which the authors construe as evidence for a disc
locking scenario. \cite{scholz_2004} also find a $mass\propto
period$ relationship extending into the BD mass regime, this faster
rotation in younger mass objects is taken as evidence for a decreasing
effectiveness of the angular momentum removal of the disc locking
mechanism, i.e. imperfect disc locking. In addition,
\cite{scholz_2004} argue that when compared with data from
younger clusters disc lifetimes can be seen to decrease with central
object mass.

\cite{scholz_2005} study rotation period data for stars in the
vicinity of $\epsilon$ Ori, with an assumed age of $\approx$3 Myrs
\citep{osorio_2002} and the ONC at an age of $\approx$2 Myrs
\citep{mayne_2008}. Rotation periods, from photometric
variability, in the range 4.7$-$87.6 hours ($\approx$ 0.2$-$3.65 days)
for BD mass stars are found. Again, \cite{scholz_2005} find
some BD objects with irregular variability indicative of accretion hot
spots on the stellar surface. \cite{scholz_2005} then analyse
the mass-period relationship of these data in comparison with data
from $\sigma$ Ori \citep[3 Myr,][]{scholz_2004}, NGC2264
\citep[2 Myr,][]{lamm_2004} and then ONC \citep[1
Myr][]{herbst_2002}. They conclude that either the $\epsilon$
Ori sample has an age of around 2$-$3 Myrs or there is still
significant disc locking (concluded using the slope of the mass-period
relation). More specifically for the BD mass regime, they find
\citep[in agreement with][]{scholz_2004} an age independent BD
lower period limit of 2$-$4 hours, therefore at any age a fraction of
BD stars rotate at short and constant period. Given that these objects
should spin up as they hydrostatically contract they invoke rotational
braking in the BD mass regime as a mechanism of removing significant
angular momentum.

\cite{joergens_2003} study the rotational periods of BD (and
very low mass stars) in the Chameleon I region. This region is $\le$ 1
Myrs old, and the authors find rotation periods of 2.19, 3.376 and
3.21 days for their BD mass counterparts. Using supplementary data
from the literature \cite{joergens_2003} propose that a shorter
lifetime of $\approx$5 Myrs for accretions discs around BDs is more
probable and is inferred from the derivation of a shorter locking
timescale. However, as discussed in the previous studies, an imperfect
disc locking mechanisms is also hypothesised as responsible for the
less significant loss in angular momentum out to ages of 10 Myrs. The
data mentioned above and other datasets are summarised and discussed
in \cite{2007prpl.conf..297H}, where a strong argument for the
mass-period relationship can be found, shown in their Fig 6, alongside
further support for either a shorter disc lifetime or imperfect disc
locking mechanism for BD stars.

Therefore, in summary, it is currently unclear if any disc locking
mechanisms efficiency or disc lifetimes are a function of mass, with
an associated phase change over the TTS to BD mass limit. Rotation
rate and disc presence data are being used to explore issue.
Therefore, to create a set of useful models to help contextualise the
observational constraints we must adopt a realistic, and bounding,
range of rotation rates. For our model grid, and associated age range
($<$10 Myrs), we have selected 0.5 and 5 days.  The lower limit is
placed at around the approximate median for the faster rotating
samples, providing a useful benchmark for exploring possible detection
limits of faster rotators with discs. The longer period is set at the
approximate edge of the slow rotating tail, providing a limiting case
for the slower rotating, disc-locked, candidates.

@@

As discussed in Section \ref{par_rotation} evidence for irregular
periodic variability has been found in BD stars with detections of
associated stellar discs. This is construed as evidence for accretion
hot spots formed as magnetically channeled material hits the stellar
surface \citep[see discussion
in][]{bouvier_1995,2007prpl.conf..297H}. The irregularity is
thought to be caused by changes in the magnetospheric structure and
accretion rate \citep{bouvier_1995}. For our model we have
assumed that disc material is disrupted at the co-rotation radius and
channeled onto the star in the form of accretion hot spots with a
characteristic temperature. Therefore, to calculate the characteristic
temperature and the resulting blackbody accretion flux we must adopt
an accretion rate and areal coverage of the accretion stream.

Little observational evidence can be found for approximate sizes of
accretion hot spots due to their more transient nature and often
smaller coverages, when compared to cooler or `plage' spots
\citep{2007prpl.conf..297H}. \cite{bouvier_1995} modeled the
size of the cool spots on solar-type stars for a selection of
periodically variable candidates. They found projections of cooler
spots, onto the stellar disc, of a few to $~$60\%.
\cite{bouvier_1995} also found projected sizes, onto the
stellar disc, of typically a few \% to around 10\% for hot spots.
\cite{bertout_1996} used observations of YY Orionis monitoring
flux amplitude variations as a function of wavelength to derive a
probable hot spot area of around 10\%. The spot temperature was also
modeled for YY Orionis in \cite{bertout_1996}, resulting in a
best fitting areal coverage of 11\%. Therefore, to bound the probable
areal coverage range of the accretion hot spots we have adopted areal
coverages of 1 and 10\%.

@@

The initial density structure has the the disc surface density
conserved, $\Sigma ^{\beta - \alpha}$. The initial scaleheight,
$h\propto r^{\beta}$, where r is the radial distance from the star,
and the density, $\rho \propto r^{-\alpha}$. The initial values of
$\alpha$ and $\beta$ were 2.1 and 1.1 respectively \citep[see][for
  full discussion of disc structure]{walker_2004}. The values
for $\alpha$ and $\beta$ were chosen to optimise resolution of the
vertically evolving disc, but minor variations are largely
inconsequential as the systems evolves from this state. In our models
we have, however, placed at the inner disc edge at the co-rotation
radius, as opposed to the dust destruction radius used in
\cite{walker_2004}. The initial disc scaleheight at 100\,AU,
$h(100)$, was set to 25\,AU. As the simulation used vertical
hydrostatic equilibrium and dust sublimation, both the disc
scaleheight and inner edge location then evolved in the systems
dependent on the input parameters. In this section we discuss the
structure of the discs in terms of these two generated
characteristics, i.e the disc scaleheight and inner edge location.

@@

\begin{figure}
  \vspace*{174pt}
  \special{psfile="Fig/inner_temp.ps"
   hoffset=235 voffset=0 hscale=30 vscale=30 angle=90}
 \caption{Figure showing data for the inner edge location ($R_{\rm
     inner}$, $R_*$). The typical accretors are shown in the
   \textit{left panel} and the extreme accretors in the \textit{right
     panel} (see text for explanation). In \textit{both panels}
   separate the systems by period, with those systems with rotation
   periods of 0.5 and 5 days shown as red and blue symbols
   respectively. In addition systems where the total change in the
   inner radius is greater than 1$R_*$ are marked as crosses (this is
   only achieved by some systems classed as extreme
   accretors).\label{inner_temp}}
\end{figure}

@@

In a future publication we will release a fitting tool to be used in
conjunction with the online database of naked and BDD systems (for
both photometry and SED fitting). Therefore, we will be performing
fits to observed systems using our grid. In this work we simply pick
an example star and highlight its position in CMD and CoCoD space,
compared to models covering the expected range of parameters.
\cite{natta_2004} derive an accretion rate of -9
log$(\frac{\dot{M}}{M_{\odot}}yr^{-1})$ for $\rho$ Ophiucus 102. For
this object \cite{natta_2002} find a spectral type of M6,
$T_{\rm eff}$ of 2700\,K, an extinction of $A_V$=3.0, a mass of
$M_*$0.04-0.08$M_{\odot}$ and a distance of 150pc. Photometry for
$\rho$ Ophiucus 102 is also published in \cite{natta_2006},
in $J$, $H$ and $K$. Additionally, $R$ and $I$ photometry has been
published for this object in \cite{wilking_2005}. We have
adjusted our model spectra in the range $M_*$=0.04-0.08$M_{\odot}$,
with accretion rates of -9 log$(\frac{\dot{M}}{M_{\odot}}yr^{-1})$
(for all remaining input parameters) to the correct distance and
reddening. Figure \ref{rho_oph_102} shows, as dots, our BDD systems
with the parameters in the range quoted, with $\rho$ Ophiucus 102 is
shown as a black star. The \textit{panels} of Figure \ref{rho_oph_102}
then show example CMDs and CoCoDs. The \textit{top left panel} shows a
$M_J$, $(J-H)_0$ CMD, the \textit{bottom left panel} a $M_R$,
$(R-I)_0$ CMD, the \textit{top right panel} a $M_J$, $(J-K)_0$ CMD and
finally, the \textit{bottom right panel} a $(J-K)_0$, $(J-H)_0$ CoCoD.

\begin{figure*}
  \vspace*{348pt}
  \special{psfile="Fig/rho_oph_102.ps"
   hoffset=470 voffset=0 hscale=60 vscale=60 angle=90}
 \caption{Figure showing CMDs and CoCoDs of systems with matching
   parameters to $\rho$ Ophiucus 102 from
   \citet{natta_2002}. The object $\rho$ Ophiucus 102 is
   shown in \textit{all panels} as a black star, the BDD systems are
   the separated by age, with the 1 and 10 Myr systems appearing as
   blue and red dots respectively. The \textit{top left panel} shows a
   $M_J$, $(J-H)_0$ CMD, the \textit{top right panel} a $M_J$,
   $(J-K)_0$ CMD, the \textit{bottom left panel} a $M_R$, $(R-I)_0$
   CMD and the \textit{bottom right panel} shows a $(J-H)_0$,
   $(J-K)_0$ CoCoD.\label{rho_oph_102}}
\end{figure*}

>>>>

Data for rotation rates, from periodic variability surveys, are widely
available for a range of different age clusters of TTS. However, fewer
studies exist on the rotation rates of BD mass objects. Rotation
period data for $\sigma$ Ori, at an age of $\approx$ 3 Myrs
\citep{mayne_2008}, is studied in \cite{scholz_2004}, where periods
are found over the range 5.78$-$74.4 hours ($\approx$ 0.24$-$3.1 days)
for BD mass objects. \cite{scholz_2005} study rotation period data for
stars in the vicinity of $\epsilon$ Ori, with an assumed age of
$\approx$3 Myrs \citep{osorio_2002} and the ONC at an age of
$\approx$2 Myrs \citep{mayne_2008}. Rotation periods, from photometric
variability, in the range 4.7$-$87.6 hours ($\approx$ 0.2$-$3.65 days)
for BD mass stars are found. \cite{joergens_2003} study the rotational
periods of BD (and very low mass stars) in the Chameleon I
region. This region is $\le$ 1 Myrs old, and the authors find rotation
periods of 2.19, 3.376 and 3.21 days for their BD mass counterparts.

In some cases the periodic variability is irregular and assumed to
come from active accretion hot spots on the BD surface \citep[see][for
discussion of variability causes]{bouvier_1995,herbst_2007},
indicative of active accretion. All the studies mentioned infer a disc
locking mechanism. Furthermore, \cite{scholz_2004} and
\cite{scholz_2005} find evidence for a $mass\propto period$
relationship extending into the BD regime. Additionally,
\cite{joergens_2003} propose a shorter lifetime of $\approx$5 Myrs for
BD discs, inferred from a shorter derived disc locking
timescale. However, as discussed in the previous studies, an imperfect
disc locking mechanisms is also hypothesised as responsible for the
less significant loss in angular momentum out to ages of 10 Myrs, for
BD discs. The data on BD rotation rates, disc presence and disc
locking are summarised and discussed in \cite{herbst_2007}.

Therefore, to create a set of useful models to help contextualise the
observational constraints, for study of disc locking mechanisms, we
must adopt a realistic, and bounding range of rotation rates. For our
model grid, and associated age range ($<$10 Myrs), we have selected
0.5 and 5 days. With the limits set at at the approximate
median of faster rotators and the edge of the slower rotators.

@@

As discussed in Section \ref{par_rotation} evidence for irregular
periodic variability has been found in BD stars with detections of
associated stellar discs. This is construed as evidence for accretion
hot spots formed as magnetically channeled material hits the stellar
surface \citep[see discussion
in][]{bouvier_1995,herbst_2007}. The irregularity is
thought to be caused by changes in the magnetospheric structure and
accretion rate \citep{bouvier_1995}. For our model we have
assumed that disc material is disrupted at the co-rotation radius and
channeled onto the star in the form of accretion hot spots with a
characteristic temperature. Therefore, to calculate the characteristic
temperature and the resulting blackbody accretion flux we must adopt
an accretion rate and areal coverage of the accretion stream.

Little observational evidence can be found for approximate sizes of
accretion hot spots due to their more transient nature and often
smaller coverages, when compared to cooler or `plage' spots
\citep{herbst_2007}. \cite{bouvier_1995} modeled the
size of the cool spots on solar-type stars for a selection of
periodically variable candidates. They found projections of cooler
spots, onto the stellar disc, of a few to $~$60\%.
\cite{bouvier_1995} also found projected sizes, onto the
stellar disc, of typically a few \% to around 10\% for hot spots.
\cite{bertout_1996} used observations of YY Orionis monitoring
flux amplitude variations as a function of wavelength to derive a
probable hot spot area of around 10\%. The spot temperature was also
modeled for YY Orionis in \cite{bertout_1996}, resulting in a
best fitting areal coverage of 11\%. Therefore, to bound the probable
areal coverage range of the accretion hot spots we have adopted areal
coverages of 1 and 10\%.

@@

The initial density structure has the the disc surface density
conserved, $\Sigma ^{\beta - \alpha}$. The initial scaleheight,
$h\propto r^{\beta}$, where r is the radial distance from the star,
and the density, $\rho \propto r^{-\alpha}$. The initial values of
$\alpha$ and $\beta$ were 2.1 and 1.1 respectively. The values
for $\alpha$ and $\beta$ were chosen to optimise resolution of the
vertically evolving disc, but minor variations are largely
inconsequential as the systems evolves from this state. In our models
we have, however, placed at the inner disc edge at the co-rotation
radius, as opposed to the dust destruction radius used in
\cite{walker_2004}. The initial disc scaleheight at 100\,AU,
$h(100)$, was set to 25\,AU. As the simulation used vertical
hydrostatic equilibrium and dust sublimation, both the disc
scaleheight and inner edge location then evolved in the systems
dependent on the input parameters. In this section we discuss the
structure of the discs in terms of these two generated
characteristics, i.e the disc scaleheight and inner edge location.

@@

\begin{figure*}
  \vspace*{348pt}
  \special{psfile="Fig/inner_temp.ps"
     hoffset=470 voffset=0 hscale=60 vscale=60 angle=90}
 \caption{Figure showing data for the inner edge location ($R_{\rm
     inner}$, $R_*$). The typical accretors are shown in the
   \textit{left panel} and the extreme accretors in the \textit{right
     panel} (see text for explanation). In \textit{both panels}
   separate the systems by period, with those systems with rotation
   periods of 0.5 and 5 days shown as red and blue symbols
   respectively. In addition systems where the total change in the
   inner radius is greater than 1$R_*$ are marked as crosses (this is
   only achieved by some systems classed as extreme
   accretors).\label{inner_temp}}
\end{figure*}

@@

In a future publication we will release a fitting tool to be used in
conjunction with the online database of naked and BDD systems (for
both photometry and SED fitting). Therefore, we will be performing
fits to observed systems using our grid. In this work we simply pick
an example star and highlight its position in CMD and CoCoD space,
compared to models covering the expected range of parameters.
\cite{natta_2004} derive an accretion rate of -9
log$(\frac{\dot{M}}{M_{\odot}}yr^{-1})$ for $\rho$ Ophiucus 102. For
this object \cite{natta_2002} find a spectral type of M6,
$T_{\rm eff}$ of 2700\,K, an extinction of $A_V$=3.0, a mass of
$M_*$0.04-0.08$M_{\odot}$ and a distance of 150pc. Photometry for
$\rho$ Ophiucus 102 is also published in \cite{natta_2006},
in $J$, $H$ and $K$. Additionally, $R$ and $I$ photometry has been
published for this object in \cite{wilking_2005}. We have
adjusted our model spectra in the range $M_*$=0.04-0.08$M_{\odot}$,
with accretion rates of -9 log$(\frac{\dot{M}}{M_{\odot}}yr^{-1})$
(for all remaining input parameters) to the correct distance and
reddening. Figure \ref{rho_oph_102} shows, as dots, our BDD systems
with the parameters in the range quoted, with $\rho$ Ophiucus 102 is
shown as a black star. The \textit{panels} of Figure \ref{rho_oph_102}
then show example CMDs and CoCoDs. The \textit{top left panel} shows a
$J$, $J-H$ CMD, the \textit{bottom left panel} a $R$,
$R-I$ CMD, the \textit{top right panel} a $J$, $J-K$ CMD and
finally, the \textit{bottom right panel} a $J-K$, $J-H$ CoCoD.

\begin{figure*}
  \vspace*{348pt}
  \special{psfile="Fig/rho_oph_102.ps"
   hoffset=470 voffset=0 hscale=60 vscale=60 angle=90}
 \caption{Figure showing CMDs and CoCoDs of systems with matching
   parameters to $\rho$ Ophiucus 102 from
   \citet{natta_2002}. The object $\rho$ Ophiucus 102 is
   shown in \textit{all panels} as a black star, the BDD systems are
   the separated by age, with the 1 and 10 Myr systems appearing as
   blue and red dots respectively. The \textit{top left panel} shows a
   $J$, $J-H$ CMD, the \textit{top right panel} a $J$,
   $J-K$ CMD, the \textit{bottom left panel} a $R$, $R-I$
   CMD and the \textit{bottom right panel} shows a $J-H$,
   $J-K$ CoCoD.\label{rho_oph_102}}
\end{figure*}
